Evolution of pre-copulatory and post-copulatory inbreeding behaviour

### 2016-09-22 ms_pre_post.Rmd
New abstract and tweaks to Introduction

Inbreeding avoidance has been widely observed through pre-copulatory and post-copulatory mechanisms as a consequence of inbreeding depression in offspring viability. However, no models compare evolution of pre-copulatory and post-copulatory inbreeding avoidance jointly, and given biologically relevant costs of, and constraints on, inbreeding strategies. Critically, evolution of post-copulatory inbreeding avoidance is restricted to females that mate multiply, potentially leading to complex evolutionary dynamics among inbreeding strategies and polyandry. Accordingly, we use individual-based modelling to track evolution of polyandry and inbreeding strategies under multiple costs and constraints to address four questions: (1) Will evolution of post-copulatory inbreeding avoidance strengthen selection for polyandry? (2) Will selection for one costly inbreeding strategy be negligible given a non-costly complementary strategy? (3) Will constraints on evolution of one inbreeding strategy increase selection on an alternative strategy? And (4) will selection for one inbreeding strategy be negligible given existence of an already adaptive alternative strategy? We show that evolution of polyandry is greatly facilitated by evolution of post-copulatory inbreeding avoidance. Costs and constraints on pre-copulatory inbreeding avoidance generally affected evolution of post-copulatory inbreeding avoidance, and vice versa. Our model thereby introduces a novel framework for predicting evolution of both pre-copulatory and post-copulatory inbreeding strategies in empirical systems.

@@

<!--- I fear we're starting down a rabbit hole here. ---> 
Severe inbreeding depression might cause selection for alleles underlying active pre- and post-copulatory inbreeding avoidance and associated polyandry, but if evolution of one strategy renders its complement redundant for avoiding inbreeding, then selection might instead be sustained over generations for only one strategy. Empirical studies of multiple species have found evidence for female pre-copulatory inbreeding avoidance [e.g., @Lihoreau2007; @Kuriwada2011; @Kingma2013; @Fischer2015], and for post-copulatory inbreeding avoidance [e.g., @Simmons2006; @Bretman2009]. Further, evidence of both pre-copulatory and post-copulatory inbreeding avoidance has been found for a handful of well-studied species, including Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015], while @Liu2014 found evidence for pre-copulatory, but not post-copulatory, inbreeding avoidance in an experimental study of cabbage beetles (*Colaphellus bowringi*). These empirical studies demonstrate that all combinations of adaptive pre-copulatory and post-copulatory inbreeding avoidance are possible, but no conceptual framework exists to predict *a priori* what combinations should be favoured by selection when both pre- and post-copulatory inbreeding avoidance, and polyandry, can evolve. Frequencies of alleles underlying active pre- and post-copulatory inbreeding avoidance and polyandry are likely to be affected by direct negative selection on phenotypes (i.e., costs). For example, females that are genetically predisposed to polyandry often pay a cost associated with searching or making themselves available to mates, thereby increasing their own risk of predation [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice (e.g., reluctance to mate with relatives) might pay a cost of increased risk of mating failure, or harm caused by sexual conflict [@Rowe1994; @Kokko2013]. Females that express post-copulatory choice might pay an energetic cost associated with physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]. If the relative costs affecting pre- and post-copulatory inbreeding avoidance differ, then alleles underlying the less costly inbreeding strategy might become fixed in the population over generations, while alleles underlying the more costly inbreeding strategy might become extinct if their effects are both costly and made redundant by the less costly strategy.

In addition to the relative importance of costs, evolution of pre- or post-copulatory inbreeding avoidance might be precluded if inbreeding avoidance is already acheived through a complementary strategy. Initial evolution of pre- and post-copulatory inbreeding avoidance strategies is unlikely to be simultaneous, so to predict evolutionary outcomes, it is necessary to consider evolution when one inbreeding avoidance strategy is initially rare, and the other strategy is common. For example, if pre-adaptation or rapid selection resulted in the fixation of alleles underlying adaptive pre-copulatory inbreeding avoidance, then alleles underlying post-copulatory inbreeding avoidance might be precluded from invading a population in which post-copulatory choice is not already expressed because the effect that such invading alleles had on decreasing total inbreeding avoidance would be weak or nelgigible.

@@

Here we apply a model of female and male reproductive interactions and associated inbreeding depression in offspring viability to quantify evolution of three separate phenotypes including female tendency for polyandry, female pre-copulatory inbreeding strategy (mating strategy), and female post-copulatory inbreeding strategy (fertilisation strategy). We use individual-based modelling and *in silico* experiments to test the hypothesis that, given severe inbreeding depression in offspring viability, evolution of inbreeding avoidance through fertilisation bias causes selection for evolution of polyandry. Additionally, we test the hypothesis that costs associated with polyandry and inbreeding strategies will strongly affect evolution of each such that inbreeding avoidance through mate choice but not fertilisation bias will evolve when inbreeding mating strategy is less costly than the combined cost of inbreeding fertilisation strategy and tendency for polyandry, while inbreeding avoidance through fertilisation bias but not mate choice will evolve when inbreeding mating strategy is more costly than the combined costs of inbreeding fertilisation strategy and tendency for polyandry. Finally, we test the hypothesis that constraints on evolution of one inbreeding strategy (e.g., inbreeding mating strategy) will increase the strength of selection on evolution of a complementary inbreeding strategy (e.g., inbreeding fertilisation strategy), and that evolution of one inbreeding strategy will be precluded if it is introduced into a population that is already fixed for a complementary inbreeding strategy that is adaptive. We thereby provide a framework for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance across reproductive systems with different relative costs and strong inbreeding depression in offspring viability.

>>>>

Inbreeding avoidance has been widely observed through pre-copulatory and post-copulatory mechanisms as a consequence of inbreeding depression in offspring viability. However, no conceptual framework compares evolution of pre-copulatory and post-copulatory inbreeding avoidance jointly, and given biologically relevant costs of inbreeding strategies. Critically, evolution of post-copulatory inbreeding avoidance is restricted to females that mate multiply, potentially leading to complex evolutionary dynamics among inbreeding strategies and polyandry. Accordingly, we use individual-based modelling to track evolution of polyandry and inbreeding strategies under multiple costs and constraints to address three questions: Will evolution of post-copulatory inbreeding avoidance strengthen selection for polyandry? Will selection for one costly inbreeding strategy be negligible given a non-costly complementary strategy? And will selection for one inbreeding strategy be negligible given existence of an already adaptive alternative strategy? We show that evolution of polyandry is greatly facilitated by post-copulatory inbreeding avoidance. Strong selection for relatively low cost inbreeding avoidance strategies was observed, while selection for functionally redundant high cost strategies decreased over generations. Finally, existence of an adaptive inbreeding strategy precluded evolution of post-copulatory, but not pre-copulatory, inbreeding avoidance. Our model thereby introduces a novel framework for predicting evolution of both pre-copulatory and post-copulatory inbreeding strategies in empirical systems.

@@

<!--- I fear we're starting down a rabbit hole here. ---> 
Severe inbreeding depression might cause selection for alleles underlying active pre- and post-copulatory inbreeding avoidance and associated polyandry, but if evolution of one strategy renders its complement redundant for avoiding inbreeding, then selection might instead be sustained over generations for only one strategy. Empirical studies of multiple species have found evidence for female pre-copulatory inbreeding avoidance [e.g., @Lihoreau2007; @Kuriwada2011; @Kingma2013; @Fischer2015], and for post-copulatory inbreeding avoidance [e.g., @Simmons2006; @Bretman2009]. Further, evidence of both pre-copulatory and post-copulatory inbreeding avoidance has been found for a handful of well-studied species, including Trinidadian guppies [*Poecilia reticulata*; @Gasparini2011; @Daniel2015] and house mice [*Mus domesticus*; @Potts1991; @Firman2015], while @Liu2014 found evidence for pre-copulatory, but not post-copulatory, inbreeding avoidance in an experimental study of cabbage beetles (*Colaphellus bowringi*). These empirical studies demonstrate that all combinations of adaptive pre-copulatory and post-copulatory inbreeding avoidance are possible, but no conceptual framework exists to predict *a priori* what combinations should be favoured by selection when both pre- and post-copulatory inbreeding avoidance, and polyandry, can evolve. Frequencies of alleles underlying active pre- and post-copulatory inbreeding avoidance and polyandry are likely to be affected by direct negative selection on phenotypes (i.e., costs). For example, females that are genetically predisposed to polyandry often pay a cost associated with searching or making themselves available to mates, thereby increasing their own risk of predation [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice (e.g., reluctance to mate with relatives) might pay a cost of increased risk of mating failure, or harm caused by sexual conflict [@Rowe1994; @Kokko2013]. Females that express post-copulatory choice might pay an energetic cost associated with physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]. If the relative costs affecting pre- and post-copulatory inbreeding avoidance differ, then alleles underlying the less costly inbreeding strategy might become fixed in the population over generations, while alleles underlying the more costly inbreeding strategy might become extinct if their effects are both costly and made redundant by the less costly strategy.

In addition to the relative importance of costs, evolution of pre- or post-copulatory inbreeding avoidance might be precluded if inbreeding avoidance is already acheived through a complementary strategy. Previous models considered the fate of a rare allele underlying inbreeding avoidance in a population initially fixed for random mating [e.g., Duthie and Reid *In press*; @Parker1979; @Parker2006; @Duthie2016]. Such models thereby isolate the invasion fitness of a single focal strategy, but when multiple strategies affect realised inbreeding, it cannot be assumed that both will invade simultaneously, nor that invasion fitness of one inbreeding strategy will be independent of another. For example, if pre-adaptation or rapid selection resulted in the fixation of alleles underlying adaptive pre-copulatory inbreeding avoidance, then alleles underlying post-copulatory inbreeding avoidance might be precluded from invading a population in which post-copulatory choice is not already expressed because the effect that such invading alleles had on decreasing total inbreeding avoidance would be weak or negligible. To predict evolutionary outcomes, it is therefore necessary to consider evolution when one inbreeding avoidance strategy is initially rare, and the other strategy is common.

<!--- Rubbish: is unlikely to be simultaneous. Testing whether or not initially rare alleles underlying inbreeding avoidance can successfully invade a population Even if invasion fitness of alleles underlying inbreeding avoidance are high when mating and fertilisation is random with respect to relatedness, initial invasion of Initial evolution of pre- and post-copulatory inbreeding avoidance strategies is unlikely to be simultaneous, so to predict evolutionary outcomes, it is necessary to consider evolution when one inbreeding avoidance strategy is initially rare, and the other strategy is common. For example, if pre-adaptation or rapid selection resulted in the fixation of alleles underlying adaptive pre-copulatory inbreeding avoidance, then alleles underlying post-copulatory inbreeding avoidance might be precluded from invading a population in which post-copulatory choice is not already expressed because the effect that such invading alleles had on decreasing total inbreeding avoidance would be weak or negligible.  --->

@@

Here we apply a model of female and male reproductive interactions and associated inbreeding depression in offspring viability to quantify evolution of three separate phenotypes including female tendency for polyandry, female pre-copulatory inbreeding strategy (mating strategy), and female post-copulatory inbreeding strategy (fertilisation strategy). We use individual-based modelling and *in silico* experiments to test the hypothesis that, given severe inbreeding depression in offspring viability, evolution of inbreeding avoidance through fertilisation bias causes selection for evolution of polyandry. Additionally, we test the hypothesis that costs associated with polyandry and inbreeding strategies will strongly affect evolution of each such that inbreeding avoidance through mate choice but not fertilisation bias will evolve when inbreeding mating strategy is less costly than the combined cost of inbreeding fertilisation strategy and tendency for polyandry, while inbreeding avoidance through fertilisation bias but not mate choice will evolve when inbreeding mating strategy is more costly than the combined costs of inbreeding fertilisation strategy and tendency for polyandry. Finally, we test the hypothesis that evolution of one inbreeding strategy will be precluded if it is introduced into a population that is already fixed for a complementary inbreeding strategy that is adaptive. We thereby provide a framework for predicting evolution of pre-copulatory and post-copulatory inbreeding avoidance across reproductive systems with different relative costs and strong inbreeding depression in offspring viability.

<!--- that constraints on evolution of one inbreeding strategy (e.g., inbreeding mating strategy) will increase the strength of selection on evolution of a complementary inbreeding strategy (e.g., inbreeding fertilisation strategy), and that --->
